Introduction to Markov state modeling with the PyEMMA software [Article v\versionnumber]

### 2018-11-29 manuscript/manuscript.tex
some edits

PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models~\cite{schuette-msm,singhal-msm-naming,noe2007jcp,chodera2007jcp,buchete-msm-2008} (MSMs).
The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.

@@

In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see Ref.~\cite{dror2012biomolecular} for a review on the MD simulation of biomolecules, and Ref.~\cite{mdtutorial} for a tutorial on MD simulations).

We further assume that the reader is familiar with the basic ideas and theory underlying Markov modeling and will only give a brief reminder of the basic concepts in Section~\ref{sec:prereq}.

For those seeking further resources, the recent perspective ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a timeline of methods advances with relevant citations,
while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology and provides the underlying mathematics in detail.
Additionally, two textbooks have been published that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on TICA~\cite{tica,tica3,kinetic-maps,tica2},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,jhp-spectral-rate-theory,bhmm-preprint},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
In its standard formulation, the creation of an MSM involves decomposing the phase or configuration space occupied by a dynamical system into a set of disjoint, discrete states,
and a transition matrix $\mathbf{P}(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
Let us make two remarks to avoid common misconceptions:
\begin{enumerate}
\item Equilibrium:
While most analysis techniques require simulation trajectories to be long enough to sample from the equilibrium distribution, this is not required for MSMs.
Because MSMs are using the \emph{conditional} probability $p_{ij}(\tau)$,
they are useful for the analysis of short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
\item Markovianity:
An MSM is a memoryless model.
Early MSM papers have argued that accurate MSMs can be found if a few states with high barriers are captured by the MSM states so as to achieve a Mori-Zwanzig projection with fast-decaying memory~\cite{swope-its,noe2007jcp,chodera2007jcp}.
The modern view, however, is that MSMs can be highly accurate if the MSM states discretize the collective coordinates of the slowest processes well~\cite{msm-jhp}.
This mainly requires that the system is characterized by only a few slow processes at lag time $\tau$,
which is true for cooperative systems such as most proteins, but not for highly frustrated systems such as glasses.
\end{enumerate}

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix $\mathbf{P}$, which is defined for the specified lag time.
For MD simulations in equilibrium, $\mathbf{P}$ should obey detailed balance which is enforced by constraining the estimation of $\mathbf{P}$ to the following equations:
\begin{equation}
\label{eq:balance}
\pi_i p_{ij} = \pi_j p_{ji},
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ and $p_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.

@@

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $\mathbf{P}(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways,msm-jhp}.
The CK property for a Markovian matrix is,
\begin{equation}
\mathbf{P}(k \tau) = \mathbf{P}^k(\tau),
\end{equation}
where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$,
whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model (see Sec.~IV.F in~\cite{msm-jhp}).

@@

The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
The dynamical process themself (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
\begin{equation}
\mathbf{P}(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
\end{equation}
where the eigenvalue-eigenvector pairs are indexed in decreasing order.
The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterize the corresponding process.
The right eigenvector $\bm{\psi}_1$ is a vector consisting of~$1$'s.

@@

\begin{itemize}
	\item Featurization -- The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
	\item Dimensionality reduction -- Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,kinetic-maps} is used to transform the features into a set of slow coordinates.
	\item Clustering -- This is the step at which the state decomposition occurs.
	The features or TICs are grouped into a set of states using a clustering algorithm such as $k$-means.
	\item Transition matrix approximation -- At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
\end{itemize}

@@

A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC.
To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
\begin{eqnarray}
c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \label{eq:c0}\\
c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t, \label{eq:ct}
\end{eqnarray}
where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
By default, PyEMMA estimates (\ref{eq:c0},\ref{eq:ct}) using symmetrization~\cite{tica}.
This symmetrization induces a significant bias when using non-equilibrium data from short trajectories~\cite{hao-variational-koopman-models}.
As an alternative, the so-called Koopman reweighting estimator is available which avoids this bias,
but comes at the cost of a large variance~\cite{hao-variational-koopman-models}.

After estimating the covariance matrices, TICA solves the generalized eigenvalue problem
\begin{equation}
\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n,
\end{equation}
to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
where the pairs of eigenvalues and independent components are sorted in descending order.
A way to measure the contribution of each independent component to the kinetics
is obtained by the kinetic distance~\cite{kinetic-maps}
which assigns a cumulative variance fraction to the first $d$ independent components:
\begin{equation}
c_d = \frac{\sum_{i=2}^d \lambda_i^2(\tau)}{\textrm{TKV}},
\end{equation}
where
\begin{equation}
\textrm{TKV} = \sum\limits_{i=2}^n \lambda_i^2(\tau)
\end{equation}
is the total kinetic variance explained by all $n$ features.

@@

Note, though, that TICA requires the dynamics to be simulated at equilibrium conditions.
To use TICA with nonequilibrium MD, e.g., subject to external forces,
or simply to perform dimension reduction on short trajectory data without worrying about reweighting,
we recommend to use VAMP~\cite{vamp-preprint}.

For all these approaches,
dimensionality reduction is performed by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
onto the leading $d$ independent components $\mathbf{U}_d=\left[\mathbf{u}_1 \dots \mathbf{u}_d\right]$,
\begin{equation}
\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
\end{equation}
where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., \SI{95}{\percent}).

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

@@

An HMM estimation always yields a model with a small number of (hidden) states
where each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see notebook~07).
As the HMMs---like MSMs---approximate the full phase-space dynamics,
we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.

For an extensive discussion of details about HMM properties and the estimation algorithm in general, we suggest Ref.~\cite{hmm-tutorial}.
For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}.
A generalized extension for estimating this type of low-dimensional projection from the data is given in Ref.~\cite{wu2015projected}.

@@

We utilize Jupyter~\cite{jupyter} notebooks to show code examples along with figures and interactive widgets to display molecules.
The user can install all necessary packages in one step using the \texttt{conda} command provided by the Anaconda Python 
stack (\url{https://conda.io/miniconda.html}).
We recommend Anaconda because it resolves and installs dependencies as well as provides pre-compiled versions of common packages.
The tutorial installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.

You can install the tutorial's dependencies in a new conda environment and start the notebook server via
\begin{verbatim}
conda create -n pyemma_tutorials
conda activate pyemma_tutorials
conda install -c conda-forge pyemma_tutorials
pyemma_tutorials
\end{verbatim}
or refer to \githubrepository{} for more detailed installation and usage instructions.

@@

The underlying software stack for running the tutorial consists of:
\begin{itemize}
    \item \textbf{PyEMMA} -- MSM/HMM estimation, validation, analysis, and visualization, and its dependencies~\cite{pyemma}
    \item mdshare -- A downloader for MD data from a public server
    \item notebook -- The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
    \item matplotlib -- A plotting library~\cite{matplotlib}
    \item nglview -- Widget for active viewing of molecular structures in Jupyter environments~\cite{nglview}
\end{itemize}

@@

This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
The first notebook~00, which we will summarize in the following, showcases the entire estimation,
validation, and analysis workflow for a small example system.
The goal of this introductory notebook~00 is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
The seven subsequent notebooks~01--07 provide in-depth lessons on specific topics,
and the last notebook~08 contains guidelines on how to deal with common problems during MSM estimation.

@@

In short, the workflow (Fig.~\ref{fig:workflowchart}) for a full analysis of an MD dataset might consist of,
\begin{itemize}
	\item extracting molecular features from the raw data (01),
	\item transforming those features into a suitable, low dimensional subspace (02),
	\item discretizing the low dimensional subsets into a state decomposition (02),
	\item estimating a maximum likelihood or Bayesian MSM from the discrete trajectories and performing validation tests (03),
	\item analyzing the stationary and kinetic properties of the MSM (04),
	\item finding metastable macrostates and applying transition path theory (TPT) to identify the pathways of conformational change (05),
	\item computing expectation values for experimental observables (06), and
	\item coarse-graining the MSM using a hidden Markov model approach (07).
\end{itemize}

For the remainder of this manuscript we will walk through the first notebook~00.
In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~\SI{0.1}{\nano\second}.
We present the results obtained in this notebook,
thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
The figures that will be displayed in the following are created in the showcase notebook~00 and can be easily reproduced.

@@

\begin{figure}[bht]
\includegraphics{figure_3}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=\SI{0.5}{\nano\second}$ shows multiple minima and
(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
\label{fig:io-to-tica}
\end{figure}

In Markov state modeling, our objective is to model the slow dynamics of a molecular process.
In order to approximate the slow dynamics in a statistically efficient manner,
a lower dimensional representation of our simulation data is necessary.
However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint}
provide a systematic means to quantitatively compare multiple representations of the simulation data.
In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
In notebook~01, we present in detail how to extract features from MD datasets and how to systematically compare them.

Throughout this tutorial, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features reflecting protein structure,
we compute the (cross-validated) VAMP-2 score (notebook~00).
Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
it is important to ensure that properties that we optimize are robust as a function of lag time. 
Consequently, we compute the VAMP-2 score at several lag times (notebook~00). 
We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
We show one example of this ranking and the absolute VAMP-2 scores for lag time~\SI{0.5}{\nano\second} in Fig.~\ref{fig:io-to-tica}b. 
We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
This suggests that backbone torsions are the best of the options evaluated for MSM construction.

@@

Subsequently, we perform TICA~\cite{tica,tica3,kinetic-maps} in order to reduce the dimension from the feature space,
which typically contains many degrees of freedom,
to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
Here, performing TICA on the backbone torsions at lag time~\SI{0.5}{\nano\second} yields a four dimensional subspace using a~\SI{95}{\percent} kinetic variance cutoff
(note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample free energy projected onto the first two independent components (ICs) exhibits several minima (Fig.~\ref{fig:io-to-tica}c).
Discrete jumps between the minima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

\subsection{Analyzing the MSM}

>>>>

PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models (MSMs)~\cite{schuette-msm,singhal-msm-naming,noe2007jcp,chodera2007jcp,buchete-msm-2008}.
The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.

@@

In this tutorial, we assume that the reader is familiar with MD and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see Ref.~\cite{dror2012biomolecular} for a review on the MD simulation of biomolecules, and Ref.~\cite{mdtutorial} for a tutorial on MD simulations).

We further assume that the reader is familiar with the basic ideas and theory underlying Markov state modeling and will only give a brief reminder of the basic concepts in Sec.~\ref{sec:prereq}.

For those seeking further resources, the recent perspective ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a timeline of methods advances with relevant citations,
while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology and details the underlying mathematics.
Additionally, two textbooks have been published that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,kinetic-maps,tica2},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,jhp-spectral-rate-theory,bhmm-preprint},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high dimensional MD simulation datasets.
In its standard formulation, the creation of an MSM involves decomposing the phase or configuration space occupied by a dynamical system into a set of disjoint, discrete states,
and a transition matrix $\mathbf{P}(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
Let us make two remarks to avoid common misconceptions:
\begin{enumerate}
\item Equilibrium:
While most analysis techniques require simulation trajectories to be long enough to sample from the equilibrium distribution, this is not required for MSMs.
Because MSMs use the \emph{conditional} probability $p_{ij}(\tau)$,
they are useful for the analysis of short simulation trajectories with arbitrary starting points---see Ref.~\cite{oom-feliks} for restrictions.
\item Markovianity:
An MSM is a memoryless model.
Early MSM papers have argued that accurate MSMs can be found if a few states with high barriers are captured by the MSM states so as to achieve a Mori-Zwanzig projection with fast-decaying memory~\cite{swope-its,noe2007jcp,chodera2007jcp}.
The modern view, however, is that MSMs can be highly accurate if the MSM states discretize the collective coordinates of the slowest processes well~\cite{msm-jhp}.
This mainly requires that the system is characterized by only a few slow processes at lag time $\tau$,
which is true for cooperative systems such as most proteins, but not for highly frustrated systems such as glasses.
\end{enumerate}

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix $\mathbf{P}(\tau)$, which is defined for the specified lag time.
For MD simulations in equilibrium, $\mathbf{P}(\tau)$ should obey detailed balance which is enforced by constraining the estimation of $\mathbf{P}(\tau)$ to the following equations:
\begin{equation}
\label{eq:balance}
\pi_i p_{ij} = \pi_j p_{ji},
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ and $p_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
for systems experiencing a pulling force or an external potential (see Ref.~\cite{Koltai2018} for a recent review on nonequilibrium MSMs).
For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces, such that Eqn.~(\ref{eq:balance}) is assumed to hold.

@@

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $\mathbf{P}(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways,msm-jhp}.
The CK property for a Markovian matrix is,
\begin{equation}
\mathbf{P}(k \tau) = \mathbf{P}^k(\tau),
\end{equation}
where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$ and $k$ is an integer larger than~$1$.
The right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model (see Sec.~IV.F in Ref.~\cite{msm-jhp}).

@@

The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
The dynamical processes themselves (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
\begin{equation}
\mathbf{P}(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
\end{equation}
where the eigenvalue-eigenvector pairs are indexed in decreasing order according to the eigenvalues.
The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterize the corresponding process.
The right eigenvector $\bm{\psi}_1$ is a vector consisting of~$1$'s.

@@

\begin{itemize}
	\item Featurization: The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
	\item Dimensionality reduction: Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
	Frequently, TICA~\cite{tica,tica3,kinetic-maps} is used to transform the features into a set of slow coordinates.
	\item Clustering: This is the step at which the state decomposition occurs.
	The features or TICs are grouped into a set of states using a clustering algorithm such as $k$-means.
	\item Transition matrix approximation: At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
\end{itemize}

@@

A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC~\cite{tica}.
To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
\begin{eqnarray}
c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \label{eq:c0}\\
c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t, \label{eq:ct}
\end{eqnarray}
where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
By default, PyEMMA estimates Eqns.~(\ref{eq:c0}) and~(\ref{eq:ct}) using symmetrization~\cite{tica}.
This symmetrization induces a significant bias when using nonequilibrium data from short trajectories~\cite{hao-variational-koopman-models}.
As an alternative, the so-called Koopman reweighting estimator is available which avoids this bias,
but comes at the cost of a large variance~\cite{hao-variational-koopman-models}.

After estimating the covariance matrices, TICA solves the generalized eigenvalue problem
\begin{equation}
\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n
\end{equation}
to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
where the pairs of eigenvalues and independent components are sorted in descending order.
A way to measure the contribution of each independent component to the kinetics
is obtained by the kinetic distance~\cite{kinetic-maps}
which assigns a cumulative variance fraction to the first $d$ independent components:
\begin{equation}
c_d = \frac{\sum_{i=2}^d \lambda_i^2(\tau)}{\textrm{TKV}},
\end{equation}
where
\begin{equation}
\textrm{TKV} = \sum\limits_{i=2}^n \lambda_i^2(\tau)
\end{equation}
is the total kinetic variance explained by all $n$ features.

@@

Note, though, that TICA requires the dynamics to be simulated at equilibrium conditions.
To use TICA with nonequilibrium MD, e.g., subject to external forces,
or simply to perform dimension reduction on short trajectory data without worrying about reweighting,
we recommend to use VAMP, which does not require reversible dynamics~\cite{vamp-preprint}.

For all of these approaches,
dimensionality reduction is performed by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
onto the leading $d$ independent components $\mathbf{U}_d=\left[\mathbf{u}_1 \dots \mathbf{u}_d\right]$,
\begin{equation}
\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
\end{equation}
where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., \SI{95}{\percent}).

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
Instead, HMMs assume that there is an underlying (hidden) dynamic process that is Markovian
and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
This is a powerful principle as we know that there is indeed an underlying process that is Markovian:
our molecular dynamics trajectories.

@@

An HMM estimation always yields a model with a small number of (hidden) states
in which each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see notebook~07).
As the HMMs---like MSMs---approximate the full phase-space dynamics,
we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.

For an extensive discussion of details about HMM properties and the estimation algorithm in general, we suggest Ref.~\cite{hmm-tutorial}.
For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}.
A generalized extension for estimating this type of low dimensional projection from the data is given in Ref.~\cite{wu2015projected}.

@@

We utilize Jupyter~\cite{jupyter} notebooks to show code examples along with figures and interactive widgets to display molecules.
The user can install all necessary packages in one step using the \texttt{conda} command provided by the Anaconda Python 
stack (\url{https://conda.io/miniconda.html}).
We recommend Anaconda because it resolves and installs dependencies and provides pre-compiled versions of common packages.
The tutorial installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.

The user can install the tutorial's dependencies in a new conda environment and start the notebook server via
\begin{verbatim}
conda create -n pyemma_tutorials
conda activate pyemma_tutorials
conda install -c conda-forge pyemma_tutorials
pyemma_tutorials
\end{verbatim}
or refer to \githubrepository{} for more detailed installation and usage instructions.

@@

The underlying software stack for running the tutorial consists of:
\begin{itemize}
    \item \textbf{PyEMMA}: MSM/HMM estimation, validation, analysis, and visualization, and its dependencies~\cite{pyemma}
    \item mdshare: A downloader for MD data from a public server
    \item notebook: The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
    \item matplotlib: A plotting library~\cite{matplotlib}
    \item nglview: Widget for active viewing of molecular structures in Jupyter environments~\cite{nglview}
\end{itemize}

@@

This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
The first notebook~(00), which we will summarize in the following, showcases the entire estimation,
validation, and analysis workflow for a small example system.
The goal of this introductory notebook is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
The seven subsequent notebooks~(01--07) provide in-depth lessons on specific topics,
and the last notebook~(08) contains guidelines on how to deal with common problems during MSM estimation.

@@

In short, the workflow (Fig.~\ref{fig:workflowchart}) for a full analysis of an MD dataset might consist of
\begin{itemize}
	\item extracting molecular features from the raw data (01),
	\item transforming those features into a suitable, low dimensional subspace (02),
	\item discretizing the low dimensional subsets into a state decomposition (02),
	\item estimating a maximum likelihood or Bayesian MSM from the discrete trajectories and performing validation tests (03),
	\item analyzing the stationary and kinetic properties of the MSM (04),
	\item finding metastable macrostates and applying transition path theory (TPT) to identify the pathways of conformational change (05),
	\item computing expectation values for experimental observables (06), and
	\item coarse-graining the MSM using a hidden Markov model approach (07).
\end{itemize}

For the remainder of this manuscript we will walk through the first notebook~(00).
In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~\SI{0.1}{\nano\second}.
We present the results obtained in this notebook,
thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
The figures that will be displayed in the following sections are created in the showcase notebook~00 and can be easily reproduced.

@@

\begin{figure}[bht]
\includegraphics{figure_3}
\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=\SI{0.5}{\nano\second}$ shows multiple minima and
(d)~the time series of the first two ICs of the first trajectory show rare transitions.}
\label{fig:io-to-tica}
\end{figure}

In Markov state modeling, our objective is to model the slow dynamics of a molecular process.
In order to approximate the slow dynamics in a statistically efficient manner,
a lower dimensional representation of our simulation data is necessary.
However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
and the more general VAMP~\cite{vamp-preprint}
provide a systematic means to quantitatively compare multiple representations of the simulation data.
In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
In notebook~01, we present in detail how to extract features from MD datasets and how to systematically compare them.

Throughout this tutorial, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features reflecting protein structure,
we compute the (cross-validated) VAMP-2 score (notebook~00).
Although we cannot optimize MSM lag times with a variational score\cite{husic2017note}, such as VAMP-2,
it is important to ensure that the properties we optimize are robust as a function of lag time. 
Consequently, we compute the VAMP-2 score at several lag times (notebook~00). 
We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
We show one example of this ranking and the absolute VAMP-2 scores for lag time~\SI{0.5}{\nano\second} in Fig.~\ref{fig:io-to-tica}b. 
We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
This suggests that backbone torsions are the best of the options evaluated for MSM construction.

@@

Subsequently, we perform TICA~\cite{tica,tica3,kinetic-maps} in order to reduce the dimension from the feature space,
which typically contains many degrees of freedom,
to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
Here, performing TICA on the backbone torsions at lag time~\SI{0.5}{\nano\second} yields a four dimensional subspace using a~\SI{95}{\percent} kinetic variance cutoff
(note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample free energy projected onto the first two independent components (ICs) exhibits several minima (Fig.~\ref{fig:io-to-tica}c).
Discrete transitions between the minima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

\subsection{MSM Analysis}
